


```{r}
library(tidyverse)
library(rtracklayer)
library(plyranges)
```


# I will be using GENCODE's basic gene annotation for gene annotations and for standardization when it comes to genes.
https://www.gencodegenes.org/human/

[It contains the basic annotations and the canonical transcripts](https://bioinformatics.stackexchange.com/questions/22551/differences-between-gencode-gtf-and-ensembl-gtf-files#:~:text=GTFs%3A%20I%20recommend%20using%20the%20Basic%20gene%20annotation%20which%20contains%20the%20canonical%20transcripts%20on%20primary%20chromosomes%2C%20but%20not%20alternative%20loci%20or%20patches.%20It%20is%20what%20I%20use%20productively%2C%20and%20what%20iirc%20GENCODE%20recommends%20for%20most%20users.)

Version: Release 49 (GRCh38.p14)

Data folder: /beagle3/haky/data/gene_annotations/GENCODE_v49

Downloaded on Thur May 14 2026
## Notes
Helps: https://www.biostars.org/p/84686/
bed files are 0-indexed, the start is the TSS - 1
GTFs are 1-indexed, so the start is the TSS

## Design choices. 
I am going to be use 0-indexing so that I don't get confused in python.

```{r}
gene_annotation <- '/beagle3/haky/data/gene_annotations/GENCODE_v49/gencode.v49.basic.annotation.gtf.gz'
```

```{r}
# dt_prot_cod <- data.table::fread("/beagle3/haky/users/temi/projects/alphagenome/files/protein_coding.tsv") %>% 
#     dplyr::select(chr, start, end, gene_id, gene_name) %>% dplyr::mutate(chr = paste0('chr', chr)) %>% dplyr::filter(chr %in% paste0('chr', 1:22)) %>% dplyr::rename(chrom = chr)
```




```{r}
# filter to protein coding genes
# gtf_data[(gtf_data$gene_type == "protein_coding") & (gtf_data$type == 'gene'), ]
# gtf_data[which((gtf_data$gene_type == "protein_coding") & (gtf_data$type == 'gene')), ]

# select protein coding genes, and chromosomes 1 to 22
# anchor the 5 prime end (which is the TSS) and subtract 1 to make it 0-indexed

canonical_tx_annotations <- import(gene_annotation) %>%
    plyranges::filter(gene_type == "protein_coding" & type == 'gene') %>%
    plyranges::filter(seqnames %in% paste0('chr', 1:22)) %>% 
    plyranges::select(gene_id, gene_name) %>%
    plyranges::anchor_5p() %>%
    plyranges::mutate(width = 1) %>%
    data.table::as.data.table() %>% 
    dplyr::rename(chrom = seqnames) %>%
    dplyr::mutate(start = start - 1L) 

head(canonical_tx_annotations)
```



```{r}
canonical_tx_annotations %>%
    data.table::fwrite("/beagle3/haky/users/temi/projects/alphagenome/files/protein_coding_genes.tss.annotations.tsv", sep = '\t', col.names = T, row.names = F, quote = F)
```

```{r}
canonical_tx_annotations <- data.table::fread("/beagle3/haky/users/temi/projects/alphagenome/files/protein_coding_genes.tss.annotations.tsv")
canonical_tx_annotations %>% dplyr::filter(gene_name == 'PRKCZ')
```


```{r}
head(gtf_data)
```

## I want to select the canonical transcripts, and one transcript by gene

```{r}
canonical_tx <- gtf_data %>%
    plyranges::filter(
        type == 'transcript',
        seqnames %in% paste0('chr', 1:22),
        any(tag == 'Ensembl_canonical')
    )

```

```{r}
canonical_tx <- gtf_data %>%
    plyranges::filter(seqnames %in% paste0('chr', 1:22), type == 'transcript', gene_type == 'protein_coding', transcript_type == 'protein_coding')
                  tag == 'MANE_Select')
```


```{r}
elementMetadata(gtf_data)$tag |> unique()
```

## Convert canonical transcripts to BED

```{r}
canonical_bed <- canonical_tx %>%
    as.data.frame() %>%
    dplyr::transmute(
        chrom  = seqnames,
        start  = start - 1L,   # GRanges is 1-based; BED is 0-based
        end    = end,
        name   = transcript_id,
        score  = 0L,
        strand = strand
    )

head(canonical_bed)
```

```{r}
bed_out <- '/beagle3/haky/users/temi/projects/alphagenome/files/canonical_transcripts.bed'
data.table::fwrite(canonical_bed, bed_out, sep = '\t', col.names = FALSE)
message('Written: ', bed_out)
```



/beagle3/haky/users/tieri/miniconda3/envs/alphaGenomeEnv/bin/pip list > ./tieri_pkgs.txt
/beagle3/haky/users/temi/software/conda_envs/alphagenome-env/bin/pip list > ./temi_pkgs.txt
diff ./tieri_pkgs.txt ./temi_pkgs.txt


conda create -p /beagle3/haky/users/temi/software/conda_envs/alphagenome-env-v2 --clone /beagle3/haky/users/tieri/miniconda3/envs/alphaGenomeEnv